Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-01 rates.tex
simplify upper bound rate
tweak to lower bound spart

These estimators have the following accuracy guarantees. 
\begin{theorem}
  \label{thm:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  Assume the estimators $\hatpimin$ and $\hatgap$ described above are
  formed from a sample path of length $n$ from an ergodic and
  reversible Markov chain.
  Let $\gap>0$ denote the spectral gap and $\pimin>0$ the minimum
  stationary probability.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation}\label{eq:piminbound}
    \Abs{\hatpimin-\pimin}
    \le
    C \,
    \Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
      +
      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
    }
  \end{equation}
  and
  \begin{equation}\label{eq:gapbound}
    \Abs{\hatgap-\gap}
    \leq
    C \,
    \Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
      + \frac{\log\frac{1}{\gap}}{\gap n}  
    }
    \,.
  \end{equation}
\end{theorem}
\todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
\todod{No, not for all values of $n$.}

@@

Next, we show that in order to estimate
$\gap$ to within a constant multiplicative accuracy in a $d$ state chain,
a sequence of length $\Omega(d\log(d)/\gap)$ is required.
Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
This hold \emph{even} if $\pimin$ is with a factor of $1/2$ of the \emph{largest} possible value of $1/d$ that it can take,
i.e., when $\vpi$ is nearly uniform.
\begin{theorem}
  \label{thm:lb-gap}
  There is an absolute constant $c>0$ such that the following holds.
  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
  (0,1/2)$.
  Consider any estimator $\hatgap$ that takes as input a random sample
  path of length $n < c d\ln(d) / \bar\gamma$ from a $d$-state
  reversible Markov chain starting from any desired initial state
  distribution.
  There is an ergodic and reversible Markov chain distribution
  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
  stationary probability $\pimin \geq 1/(2d)$ such that
  \[
    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
  \]
\end{theorem}

>>>>

These estimators have the following accuracy guarantees. 
\begin{theorem}
  \label{thm:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  Assume the estimators $\hatpimin$ and $\hatgap$ described above are
  formed from a sample path of length $n$ from an ergodic and
  reversible Markov chain.
  Let $\gap>0$ denote the spectral gap and $\pimin>0$ the minimum
  stationary probability.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation}\label{eq:piminbound}
    \Abs{\hatpimin-\pimin}
    \le
    C \,
    \Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
      +
      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
    }
  \end{equation}
  and
  \begin{equation}\label{eq:gapbound}
    \Abs{\hatgap-\gap}
    \leq
    C \,
    \Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
%      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
      + \frac{\log\frac{1}{\gap}}{\gap n}  
    }
    \,.
  \end{equation}
\end{theorem}
\todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
\todod{Oh, I see what you mean---yes, that's right.}

@@

Next, we show that in order to estimate
$\gap$ to within a constant multiplicative accuracy in a $d$ state chain,
a sequence of length $\Omega(d\log(d)/\gap)$ is required.
Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
This hold \emph{even} if $\pimin$ is within a factor of two of the
\emph{largest} possible value of $1/d$ that it can take, i.e., when
$\vpi$ is nearly uniform.
\begin{theorem}
  \label{thm:lb-gap}
  There is an absolute constant $c>0$ such that the following holds.
  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
  (0,1/2)$.
  Consider any estimator $\hatgap$ that takes as input a random sample
  path of length $n < c d\ln(d) / \bar\gamma$ from a $d$-state
  reversible Markov chain starting from any desired initial state
  distribution.
  There is an ergodic and reversible Markov chain distribution
  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
  stationary probability $\pimin \geq 1/(2d)$ such that
  \[
    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
  \]
\end{theorem}
